A Plasma Torus Around a Young Low-Mass Star

### 2025-05-01 paper/ms.tex
update

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\carnegie}{Observatories of the Carnegie Institution for Science, Pasadena, CA 91101, USA}
\newcommand{\caltech}{Department of Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
%%%%%%%%%%%%%%%%

@@

\shortauthors{L.~G.~Bouma}

@@

@@

The circumstellar material -- either pure plasma or dusty plasma --
could originate either from the star or from an external source.
Plausible external sources include an undetected old disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io torus (e.g.~\citealt{Bagenal1981}),
although with a very different composition.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been previously observed to exhibit sinusoidal
time-varying Balmer emission
\citep{Donati2000,Townsend2005,Dunstone2006,Skelly2008}, similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous
transient absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma in
the solar corona that can last days to weeks \citep{VialEngvold2015}.
In the case of the Sun, these condensations fall back to the photosphere
because gravity is stronger than any magnetic tension or centrifugal
force capable of sustaining them.  However it has been understood at
least since work by \citet{Donati2000} that these ``prominence systems''
can be longitudinally extended, forming a trapped ring of plasma.  While
the issue of why these previously known spectroscopic variables do
not show complex photometric variability remains open, plausible
explanations include that they are not in the required transiting
geometry, or that they lack the necessary source of opacity.  What we
have in the case of TIC\,141146667 is a system that finally exhibits
both sets of phenomena.

@@

\begin{acknowledgements}
  We thank M.~Jardine, A.~Weinberger, B.~Tofflemire,
  J.~Spake, J.~Winn, and L.~Hillenbrand for insightful conversations
  that significantly informed this work, as well as A.~Howard and
  H.~Isaacson for their assistance reducing the HIRES spectra.

  This work was supported by the Carnegie Fellowship and by the
  Heising-Simons 51~Pegasi~b Fellowship.
  %

@@

@@

Figure~\ref{fig:hirescuts} shows a few regions of
interest in the HIRES spectra, which cover 3650-7960\,\AA.  
H$\beta$ falls in an inter-order gap and was not observed.
Higher order Balmer lines including H$\gamma$ and H$\delta$
($n=5\rightarrow2$ and $n=6\rightarrow2$) do show variability outside
the line core.  However, this variability is not clearly periodic in the
same manner as the emission seen in H$\alpha$.  This could be because
there are insufficient hydrogen ions in the relevant excited states, or
because the spectra have lower precision in these bluer regions.  The Ca
HK doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are too blended to be usable.

@@

Finally, the Li\,\textsc{I} 6708\,\AA\ doublet, which shows no obvious
absorption, as well as the broad K\,\textsc{I} 7699\,\AA\ resonance
line are both visible in Figure~\ref{fig:hirescuts}.  In
the latter, narrow telluric absorption features overlap the blue wing
of the line.  Neither of these regions shows any notable variability.

>>>>

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\carnegie}{Observatories of the Carnegie Institution for Science, Pasadena, CA 91101, USA}
\newcommand{\caltech}{Department of Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\standrews}{School of Physics and Astronomy, University of St Andrews, North Haugh, St Andrews, Fife, Scotland KY16 YSS, UK}
%%%%%%%%%%%%%%%%

@@

\shortauthors{Bouma et al.}

@@

\author[orcid=0000-0002-1466-5236,sname='Jardine']{Moira~M.~Jardine}
\affiliation{\standrews}
\email{mmj@st-andrews.ac.uk}

@@

The circumstellar material -- either pure plasma or dusty plasma --
could originate either from the star or from an external source.
Plausible external sources include an undetected old disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io torus (e.g.~\citealt{Bagenal1981}),
although with a very different composition.
% (Because the Io ring has lots of sodium and S II (6717) emission.
% Sodium is ambiguous in HIRES due to order placement, but the sulfur
% is not there.  With that said, it is much hotter in an M dwarf
% magnetosphere than around Jupiter; the ionization balance would be
% quite different.)

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been previously observed to exhibit sinusoidal
time-varying Balmer emission
\citep{Donati2000,Townsend2005b,Dunstone2006,Skelly2008}, similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous
transient absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma in
the solar corona that can last days to weeks \citep{VialEngvold2015}.
In the case of the Sun, these condensations fall back to the photosphere
because gravity is stronger than any magnetic tension or centrifugal
force capable of sustaining them.  However it has been understood at
least since work by \citet{Donati2000} that these ``prominence systems''
can be longitudinally extended, forming a trapped ring of plasma.  While
the issue of why these previously known spectroscopic variables do
not show complex photometric variability remains open, plausible
explanations include that they are not in the required transiting
geometry, or that they lack the necessary source of opacity.  What we
have in the case of TIC\,141146667 is a system that finally exhibits
both sets of phenomena.

@@

\begin{acknowledgements}
  We thank B.~Tofflemire, A.~Weinberger, 
  J.~Spake, J.~Winn, and L.~Hillenbrand for insightful conversations
  that significantly informed this work, as well as A.~Howard and
  H.~Isaacson for their assistance reducing the HIRES spectra.

  LGB acknowledges support from the Carnegie Fellowship and the
  Heising-Simons 51~Pegasi~b Fellowship.
  MMJ acknowledges support from STFC consolidated grant
  number ST/R000824/1.
  %

@@

{\it \large Contributions}: 
LGB led the inception, design, execution, and interpretation of the
research.  MMJ contributed to the interpretation.

@@

Figure~\ref{fig:hirescuts} shows a few regions of
interest in the HIRES spectra, which cover 3650-7960\,\AA.  
H$\beta$ falls in an inter-order gap and was not observed.
Higher order Balmer lines including H$\gamma$ and H$\delta$
($n=5\rightarrow2$ and $n=6\rightarrow2$) do show variability outside
the line core.  However, this variability is not clearly periodic in the
same manner as the emission seen in H$\alpha$.  This could be because
there are insufficient hydrogen atoms in the relevant excited states, or
because the spectra have lower precision in these bluer regions.  The Ca
HK doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are too blended to be usable.

@@

Finally, the Li\,\textsc{I} 6708\,\AA\ region, which shows no obvious
absorption, as well as the broad K\,\textsc{I} 7699\,\AA\ resonance
line are both visible in Figure~\ref{fig:hirescuts}.  In
the latter, narrow telluric absorption features overlap the blue wing
of the line.  Neither of these regions shows any notable variability.
